Quantifying the importance of an inference model in Bayesian phylogenetics

### 2020-03-30 content.tex
Changes @Giappo

The goal of \verb;pirouette; is to quantify the importance of a tree prior.
It does so by measuring the inference error made 
for a given reconstructed phylogeny, 
simulated under a (usually novel) diversification model.
We refer to the model that has generated the given tree
as the 'generative tree model' $\mathit{p_{G}}$.
We define the birth-death (BD) model [\cite{nee1994reconstructed}] as
the standard tree model, as many (non-standard) tree models 
have a parameter setting for which they reduce to this model. 
One such example is the diversity-dependent (DD) diversification 
model [\cite{etienne2020package, etienne2012diversity}]:
the DD model is a BD model with a speciation or extinction rate that depends 
on the number of species and a clade-level carrying capacity. For a carrying capacity
of infinity, the DD model reduces to the BD model.
When benchmarking a novel tree model, 
one will typically construct phylogenies 
for different combinations of the diversification model's parameters, 
to assess under which scenarios the inference error cannot be neglected. 
While we recommend many replicate simulations 
when assessing a novel tree prior, 
our examples contain only one replicate,
as the goal is to show the workings of \verb;pirouette;,
instead of doing an extensive analysis.
The supplementary material, however, does show the results
of replicated runs under multiple settings.

@@

\begin{figure}
  \centering
  \includegraphics[width = \textwidth]{workflow4.png}
  \caption{
    \texttt{pirouette} pipeline.
    The pipeline starts from a phylogeny (1a) simulated by the 
    generative tree model 
    $\mathit{p_{G}}$.
    The phylogeny is converted to an alignment (2a) using the generative 
    alignment model 
    $\mathit{A} = (\mathit{c_{G}}, \mathit{s_{G}})$, composed of a clock model and a site model. 
    The user defines one or more experiments.
    For each candidate experiment $\mathit{X_{i}}$ 
    (a combination of inference model $\mathit{I_{i}}$ and condition $\mathit{C_{i}}$),
    if its condition $\mathit{C_{i}}$ is 
    satisfied (which can depend on the alignment), 
    the corresponding inference model $\mathit{I} = \mathit{I_{i}}$ is selected
    to be used in the next step.
    The inference models (3a) of the selected experiments use the alignment (2a) 
    to each create a Bayesian posterior of (parameter estimates and) 
    phylogenies (4a). 
    Each of the posteriors' trees is compared to the true phylogeny (1a) 
    using the error measure $\mathit{E}$, 
    resulting in an error distribution (5a). 
    Optionally, for each selected inference model a twin pipeline can be run.
    A twin phylogeny (1b) can be generated from the original 
    phylogeny (1a) using the twin tree model $\mathit{p_{t}}$, 
    selected among standard diversification models; 
    the default option is the standard BD model, 
    with parameters estimated from the original phylogeny.
    A twin alignment (2b) is then simulated from the twin phylogeny 
    using clock model $\mathit{c_{G}}$ and site model $\mathit{s_{G}}$ 
    imported from the generative model. 
    The twin pipeline follows the procedure of the main pipeline, 
    resulting in a twin error distribution (5b).
  }
  \label{fig:pipeline}
\end{figure}

@@

We show the usage of \verb;pirouette; on an tree generated 
by the non-standard diversity-dependent (DD) tree model [\citep{etienne2020package, etienne2012diversity}],
which is a BD model with a speciation rate that is dependent on the number of species.

@@

The candidate model that had highest evidence given the simulated alignment,
was JC, RLN, BD (see Table \ref{tab:options} for the meaning of these 
abbreviations). JC was indeed the site model used when simulating the
alignments. The RLN clock model is an unexpected choice: the RLN clock
model assumes that mutation rates differ between species, whereas the
alignment was simulated with equal mutation rates for all species.
The BD model is indeed closer to the DD model than Yule, as both BD and DD
model allow for extinction, while Yule does not.

@@

As noted in the introduction, Duch\^{e}ne and colleagues [\cite{duchene2018phylodynamic}],
also developed a method to assess the adequacy of a tree model
on empirical trees. They simulate trees from the posterior distribution of the parameters and then compare this to the originally inferred tree using tree statistics, to determine whether the assumed tree model in inference indeed generates the tree as inferred. This is useful if these trees match, but when they do not, this does not mean that the inferred tree is incorrect; if sufficient data is available the species tree prior may not be important, and hence the inference may be adequate even though the assumed species tree prior is not. In short, the approach is applied to empirical trees and compares the posterior and prior distribution of trees (with the latter generated with the posterior parameters!).
\verb;pirouette; aims to identify when assuming standard priors, not only for the species tree - which was our main focus here -, but also for the site and clock model, leads to incorrect inference if one believes more complex diversification, site or clock models are operating than can be currently accommodated in inference. In short, our approach applies to simulated trees and compares the posterior distributions of trees generated with a standard and non-standard model, but inferred with a standard one. The two methods therefore complement one another.

In conclusion, \verb;pirouette; can show the errors to be expected
when the tree model assumed in inference is different from the 
actual generative tree model.
The user can then judge whether or not a new tree prior, 
tailored to the novel tree model, is needed in our Bayesian phylogenetic tool.

>>>>

The goal of \verb;pirouette; is to quantify the importance of a tree prior.
It does so by measuring the inference error made 
for a given reconstructed phylogeny, 
simulated under a (usually novel) diversification model.
We refer to the model that has generated the given tree
as the 'generative tree model' $\mathit{p_{G}}$.
We define the birth-death (BD) model [\cite{nee1994reconstructed}] as
the standard tree model, as many (non-standard) tree models 
have a parameter setting for which they reduce to this model. 
One such example is the diversity-dependent (DD) diversification 
model [\cite{DDD, etienne2012diversity}]:
the DD model is a BD model with a speciation or extinction rate that depends 
on the number of species and a clade-level carrying capacity. For a carrying capacity
of infinity, the DD model reduces to the BD model.
When benchmarking a novel tree model, 
one will typically construct phylogenies 
for different combinations of the diversification model's parameters, 
to assess under which scenarios the inference error cannot be neglected. 
While we recommend many replicate simulations 
when assessing a novel tree prior, 
our examples contain only one replicate,
as the goal is to show the workings of \verb;pirouette;,
instead of doing an extensive analysis.
The supplementary material, however, does show the results
of replicated runs under multiple settings.

@@

\begin{figure}
  \centering
  \includegraphics[width = 0.99\textwidth]{workflow4.png}
  \caption{
    \texttt{pirouette} pipeline.
    The pipeline starts from a phylogeny (1a) simulated by the 
    generative tree model 
    $\mathit{p_{G}}$.
    The phylogeny is converted to an alignment (2a) using the generative 
    alignment model 
    $\mathit{A} = (\mathit{c_{G}}, \mathit{s_{G}})$, composed of a clock model and a site model. 
    The user defines one or more experiments.
    For each candidate experiment $\mathit{X_{i}}$ 
    (a combination of inference model $\mathit{I_{i}}$ and condition $\mathit{C_{i}}$),
    if its condition $\mathit{C_{i}}$ is 
    satisfied (which can depend on the alignment), 
    the corresponding inference model $\mathit{I} = \mathit{I_{i}}$ is selected
    to be used in the next step.
    The inference models (3a) of the selected experiments use the alignment (2a) 
    to each create a Bayesian posterior of (parameter estimates and) 
    phylogenies (4a). 
    Each of the posteriors' trees is compared to the true phylogeny (1a) 
    using the error measure $\mathit{E}$, 
    resulting in an error distribution (5a). 
    Optionally, for each selected inference model a twin pipeline can be run.
    A twin phylogeny (1b) can be generated from the original 
    phylogeny (1a) using the twin tree model $\mathit{p_{t}}$, 
    selected among standard diversification models; 
    the default option is the standard BD model, 
    with parameters estimated from the original phylogeny.
    A twin alignment (2b) is then simulated from the twin phylogeny 
    using clock model $\mathit{c_{G}}$ and site model $\mathit{s_{G}}$ 
    imported from the generative model. 
    The twin pipeline follows the procedure of the main pipeline, 
    resulting in a twin error distribution (5b).
  }
  \label{fig:pipeline}
\end{figure}

@@

We show the usage of \verb;pirouette; on an tree generated 
by the non-standard diversity-dependent (DD) tree model [\citep{DDD, etienne2012diversity}],
which is a BD model with a speciation rate that is dependent on the number of species.

@@

The candidate model that had highest evidence given the simulated alignment,
was JC, RLN, BD (see Table \ref{tab:options} for the meaning of these 
abbreviations). JC was indeed the site model used when simulating the
alignments. The RLN clock model is an unexpected choice: the RLN clock
model assumes that mutation rates differ between lineages, whereas the
alignment was simulated with equal mutation rates for all lineages.
The BD model is indeed closer to the DD model than Yule, as both BD and DD
model allow for extinction, while Yule does not.

@@

As noted in the introduction, Duch\^{e}ne and colleagues [\cite{duchene2018phylodynamic}],
also developed a method to assess the adequacy of a tree model
on empirical trees. They simulate trees from the posterior distribution of the parameters and then compare this to the originally inferred tree using tree statistics, to determine whether the assumed tree model in inference indeed generates the tree as inferred. This is useful if these trees match, but when they do not, this does not mean that the inferred tree is incorrect; if sufficient data is available the species tree prior may not be important, and hence the inference may be adequate even though the assumed species tree prior is not. In short, the approach is applied to empirical trees and compares the posterior and prior distribution of trees (with the latter generated with the posterior parameters!).
\verb;pirouette; aims to identify when assuming standard priors for the species tree leads to incorrect inference if one believes more complex diversification models are operating than can be currently accommodated in inference. In short, our approach applies to simulated trees and compares the posterior distributions of trees generated with a standard and non-standard model, but inferred with a standard one. The two methods therefore complement one another.

However, we note that the \verb;pirouette; pipeline is not restricted to exploring the effects of a new species tree model. The pipeline can also be used to explore the effects of non-standard clock or site models, such as relaxed clock models with a non-standard distribution, correlated substitutions on sister lineages, or elevated substitutions rates during speciation events. It is, however, beyond the scope of this paper to discuss all these options in more detail.

In conclusion, \verb;pirouette; can show the errors to be expected
when the model assumed in inference is different from the 
actual generative model.
The user can then judge whether or not this new model should be implemented in our Bayesian phylogenetic tool.
